## Setting Up

We'll start with the `week_8_1.csv` file again today. Load it with the usual packages.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
week_8_1 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/week_8_1.csv")
```


## Warm Up

1. Test if there is a significant difference (at the 95% confidence level) in average number of political actions between respondents who self identify as lower class and respondents who self identify as working class. Use the `polactions` variable for political actions and the `class` variable for subjective class identification.

### REPLACE THIS LINE WITH YOUR CODE


2. Test if there is a significant difference (at the 95% confidence level) in average age between respondents who work for someone else and respondents who are self employed. Use the `age` variable for age and the `wrkslf` variable for work status.

### REPLACE THIS LINE WITH YOUR CODE


## Comparing Proportions

We have seen over the last couple weeks that proportions and means have different standard errors. That is still true when we are comparing differences in proportions. Also, when working with differences in proportions, we will use different SEs for confidence intervals and for significance tests.

The SE for a CI of a difference in proportions is very similar to what we used last week as the SE for a CI of a proportion:

$\Large{se = \sqrt{ \frac {\hat{\pi}_1 (1 - \hat{\pi}_1)} {n_1} + \frac {\hat{\pi}_2 (1 - \hat{\pi}_2)} {n_2} }}$

The SE for difference uses a pooled estimate of the proportion and the n's for each sample:

$\Large{se = \sqrt{ \hat{\pi}(1-\hat{\pi})  (\frac {1} {n_1} + \frac{1}{n_2})  }}$

The null hypothesis is that proportions are the same in each sample. Formally, this is written as:

$\Large{\pi_2 - \pi_1 = 0}$

The test statistic is calculated as:

$\Large{\frac{\hat{\pi}_1 - {\hat{\pi}_2}}{se}}$


## Example

Let's look at racial differences in the proportion of female respondents who have their first child when they are teenagers.

What do we need?

Start by cleaning up the data frame so we only have female respondents. You can keep the same data frame name.

### REPLACE THIS LINE WITH YOUR CODE

Let's review a proportion table with `racehisp` in the rows and `agekdbrn` in the columns.

### REPLACE THIS LINE WITH YOUR CODE

This is going to be easier if we make a binary variable called `teen_birth` where respondents who were 19 or younger when they had their first birth are coded with a 1 and everyone else gets a 0.

### REPLACE THIS LINE WITH YOUR CODE

And look at the proportions with a 1 by race.

### REPLACE THIS LINE WITH YOUR CODE

Is there a significant difference in the Black and Hispanic proportions at the .05 alpha level?

To find out, we'll use `prop.test()` again. The change in how we use `prop.test()` this week compared to last week is that we now need a full frequency table rather than two frequencies. So in this example we need a table of the number of respondents in each `racehisp` category that interests us who have a 0 and 1 for our `teen_birth` variable.

We know how to get a frequency table (not a proportion table) for *all* the racehisp categories. We can filter to only keep respondents with "Black" or "Hispanic" values for racehisp. But the "White" and "Other" values still technically exist so we have to use the `droplevels()` function to get rid of them. Then save the table we need as an object called `racehisp_teenbirth_table`.

```{r}
black_hispanic <- week_8_1 |> 
   filter(racehisp == "Black" | racehisp == "Hispanic") |> 
   droplevels() # Add this to drop the factor levels you don't want in the table
```


```{r}
racehisp_teenbirth_table <- table(black_hispanic$racehisp,
                                  black_hispanic$teen_birth)

racehisp_teenbirth_table
```

...and feed that object into `prop.test()`.

```{r}
prop.test(racehisp_teenbirth_table)
```

How do you interpret these results?

What are the `prop 1` and `prop 2` sample estimates?

```{r}
prop.table(racehisp_teenbirth_table,1)
```


# Exercises

Let's try a couple examples with a new dataset

```{r}
week_8_2 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/week_8_2.csv")
```


## Exercise 1

At the .05 level, test the difference between the proportion of working class respondents who say that hard work is how people get ahead and the proportion of middle class respondents who say that hard work is how people get ahead. Use the variable name `hardwork` when you make the binary variable for "Hard work".

### REPLACE THIS LINE WITH YOUR CODE


## Exercise 2

At the .05 level, test the difference between the proportion of Black respondents who say that hard work is how people get ahead and the proportion of white respondents who say that hard work is how people get ahead. Use the variable name `hardwork` when you make the binary variable for "Hard work".

### REPLACE THIS LINE WITH YOUR CODE