---
title: "Differences in Means"
date: "November 1, 2021"
author: "SOAN 385 - Social Statistics"
fontsize: 11pt
output:
  pdf_document: default
  html_notebook: default
---

## Setting Up

We'll use the `week_8_1.csv` file today. Load it as a data frame called `week_8_1` and load the usual packages.

```{r, include = FALSE, warning = FALSE, message = FALSE}
week_8_1 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/week_8_1.csv")
library(tidyverse)
```


## Introducing Differences in Means

This week, we move from calculating confidence intervals for means to calculating confidence intervals for *differences* in means. In this example, we only want the observations with non-missing values for the `memnum` variable. Create a new data frame called `subset` with the filtered observations.

### REPLACE THIS LINE WITH YOUR CODE

```{r filter out missing values}
subset <- filter(week_8_1, !is.na(memnum))
```

We want to compare mean memberships across two degree categories. This sounds like a job for binary variables or indexing. Let's use a binary variable first.

Create a binary variable where respondents with at least a college degree get a 1 and everyone else gets a 0:

### REPLACE THIS LINE WITH YOUR CODE

```{r create binary variable}
subset <- mutate(subset, college = 
                      ifelse(str_detect(degree, 
                                        "Degree"), 1, 0))
```

For the standard error formula, we will need the number of respondents in each category of `college`:

### REPLACE THIS LINE WITH YOUR CODE

```{r table of college}
table(subset$college)
```

Using the `college` binary variable, find the mean memberships for college degree holders and the mean memberships for non-college degree holders:

### REPLACE THIS LINE WITH YOUR CODE

```{r find means}
mean(subset$memnum[subset$college==1])
mean(subset$memnum[subset$college==0])
```

Now that we have the means for both groups, we want to know if the difference between them is significant. Eventually we will find a test statistic and a p-value for the difference. But we'll start by finding the confidence interval for the difference. The confidence interval formula is similar to other confidence interval formulas we have seen:

$\Large{CI = (\bar{y}_2 - \bar{y}_1) \pm t(se)}$

Our first step is to find the standard error of the difference:

$\Large{se = \sqrt{ \frac{var_1}{n_1} + \frac{var_2}{n_2}}}$

Start with the difference between the means:

```{r find the difference}
diff <- 2.551163 - 1.218357
diff
```

Now find the standard error:

```{r find the standard error}
diffse_college0 <- var(subset$memnum[subset$college==0]) / 
                              1035

diffse_college1 <- var(subset$memnum[subset$college==1]) / 
                              430

diffse <- sqrt(diffse_college0 + diffse_college1)

diffse
```

Then use `diff` and `diffse` to construct the 95% of the confidence interval for the difference:

### REPLACE THIS LINE WITH YOUR CODE

```{r calculate confidence intervals}
diff_ll95 <- diff - 1.96*diffse
diff_ul95 <- diff + 1.96*diffse
diff_ci95 <- c(diff_ll95, diff, diff_ul95)

diff_ci95
```


## Significance of Differences in Means

The significance test for a difference in means is similar to the significance test for a mean. We need a null hypothesis (usually that the true difference is zero), an alternative hypothesis (usually that the true difference is not zero), a test statistic, a critical value (1.96 for 95% confidence level, 2.58 for 99% confidence level if sample sizes are sufficient), a p-value, and an alpha level (.05 for 95% confidence level, .01 for 99% confidence level).

The test statistic is the difference in means divided by the standard error of that difference. For our example:

```{r calculate test statistic}

# Test Statistic:
(2.551163 - 1.218357) / diffse
```

How would you interpret this test statistic for a confidence level of 95%?


## Comparing Means in R

The `t.test()` function can be expanded to test the difference between two means. Simply include the two means you want to compare:

```{r ttest in r}
t.test(subset$memnum[subset$college==1], 
          subset$memnum[subset$college==0])
```

How would you interpret this output?

Try another example. Is the difference in mean memberships between those in the "Some College" degree category and those in the "HS Diploma" degree category significantly different at the .01 alpha level?

### REPLACE THIS LINE WITH YOUR CODE

```{r ttest in r 2}
t.test(subset$memnum[subset$degree=="Some College"],
       subset$memnum[subset$degree=="HS Diploma"],
       conf.level = .99)
```


## Group Exercises

Here are some of the other comparisons we have discussed that we can finally test:

- Age at first birth (`agekdbrn`) by two values of race (`racehisp`); only use female respondents' age at first birth
- Age (`age`) by self employment status (`wrkslf`)
- Number of political actions (`polactions`) by sex (`sex`)
- Number of political actions (`polactions`) by sex (`sex`) *and* two values of class (`class`)


Use the full `week_8_1` dataframe (not `subset`) to test these differences.

Give a sociological hypothesis for why you expect the difference you are testing will or will not be statistically significant.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
# Average age at first birth between Black and white respondents
t.test(subset$agekdbrn[subset$racehisp=="Black"],
       subset$agekdbrn[subset$racehisp=="White"])
```

```{r}
# Average age at first birth between Black and white respondents who have college degrees
t.test(subset$agekdbrn[subset$racehisp=="Black" &
                         subset$college==1],
       subset$agekdbrn[subset$racehisp=="White" &
                         subset$college==1])
```

```{r}
# Average number of political actions between male and female respondents
t.test(week_8_1$polactions[week_8_1$sex=="Male"],
       week_8_1$polactions[week_8_1$sex=="Female"])
```

```{r}
# Average age between respondents who work for someone else and who are self employed
t.test(week_8_1$age[week_8_1$wrkslf=="Work For Someone Else"],
       week_8_1$age[week_8_1$wrkslf=="Self Employed"])
```

```{r}
# Average number of political actions by sex and class
t.test(week_8_1$polactions[week_8_1$class=="Working" & week_8_1$sex=="Male"],
       week_8_1$polactions[week_8_1$class=="Working" & week_8_1$sex=="Female"],
       conf.level = .99)
```

