---
title: "Differences in Proportions"
author: "Matt Lawrence"
date: "10/30/2019"
output: pdf_document
---

## Setting Up

We'll start with the `week_8_1.csv` file again today. Load it with the usual packages.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
week_8_1 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/week_8_1.csv")
```


## Warm Up

Test if there is a significant difference (at the 95% confidence level) in average age between respondents who work for someone else and respondents who are self employed. Use the `age` variable for age and the `wrkslf` variable for work status.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
# Average age between respondents who work for someone else and who are self employed
t.test(week_8_1$age[week_8_1$wrkslf=="Work For Someone Else"],
       week_8_1$age[week_8_1$wrkslf=="Self Employed"])
```

Test if there is a significant difference (at the 95% confidence level) in average number of political actions between respondents who self identify as lower class and respondents who self identify as working class. Use the `polactions` variable for political actions and the `wrkslf` variable for work status.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
# Average number of political actions between upper and working class respondents
t.test(week_8_1$polactions[week_8_1$class=="Lower"],
       week_8_1$polactions[week_8_1$class=="Working"],
       conf.level = .95)
```

## Comparing Proportions

We have seen over the last couple weeks that proportions and means have different standard errors. That is still true when we are comparing differences in proportions. Also, when working with differences in proportions, we will use different SEs for confidence intervals and for significance tests.

The SE for a CI of a difference in proportions is very similar to what we used last week as the SE for a CI of a proportion:

$\Large{se = \sqrt{ \frac {\hat{\pi}_1 (1 - \hat{\pi}_1)} {n_1} + \frac {\hat{\pi}_2 (1 - \hat{\pi}_2)} {n_2} }}$

The SE for difference uses a pooled estimate of the proportion and the n's for each sample:

$\Large{se = \sqrt{ \hat{\pi}(1-\hat{\pi})  (\frac {1} {n_1} + \frac{1}{n_2})  }}$

The null hypothesis is that proportions are the same in each sample. Formally, this is written as:

$\Large{\pi_2 - \pi_1 = 0}$

The test statistic is calculated as:

$\Large{\frac{\hat{\pi}_1 - {\hat{\pi}_2}}{se}}$


## Example

Let's look at racial differences in the proportion of female respondents who have their first child when they are teenagers.

What do we need?

Start by cleaning up the data frame so we only have female respondents. You can keep the same data frame name.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
week_8_1 <- filter(week_8_1, sex == "Female")
```


Let's review a proportion table with `racehisp` in the rows and `agekdbrn` in the columns.

### REPLACE THIS LINE WITH YOUR CODE

```{r proportion table}
round(prop.table(table(week_8_1$racehisp, 
      week_8_1$agekdbrn),1),3)
```


This is going to be easier if we make a binary variable called `teen_birth` where respondents who were 19 or younger when they had their first birth are coded with a 1 and everyone else gets a 0.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
week_8_1 <- week_8_1 |> 
   mutate(teen_birth = ifelse(agekdbrn<=19, 1, 0))
```

And look at the proportions with a 1 by race.

### REPLACE THIS LINE WITH YOUR CODE


```{r}
week_8_1 |> 
   group_by(racehisp) |> 
   summarise(prop_teen_birth = mean(teen_birth, na.rm=TRUE))
```

Is there a significant difference in the Black and Hispanic proportions at the .05 alpha level?

To find out, we'll use `prop.test()` again. The change in how we use `prop.test()` this week compared to last week is that we now need a full frequency table rather than two frequencies. So in this example we need a table of the number of respondents in each `racehisp` category that interests us who have a 0 and 1 for our `teen_birth` variable.

We know how to get a frequency table (not a proportion table) for *all* the racehisp categories. We can filter to keep respondents with "Black" or "Hispanic" values for racehisp. But the "White" and "Other" values still technically exist so we have to use the `droplevels()` function to get rid of them. Then use that new data frame to save the table we need as an object called `racehisp_teenbirth_table`.

```{r}
black_hispanic <- week_8_1 |> 
   filter(racehisp == "Black" | racehisp == "Hispanic") |> 
   droplevels() # Add this to drop the factor levels you don't want in the table
```


```{r}
racehisp_teenbirth_table <- table(black_hispanic$racehisp,
                                  black_hispanic$teen_birth)

racehisp_teenbirth_table
```

...and feed that object into `prop.test()`.

```{r}
prop.test(racehisp_teenbirth_table)
```

How do you interpret these results?

What are the `prop 1` and `prop 2` sample estimates?

```{r}
prop.table(racehisp_teenbirth_table,1)
```


# Exercises

Let's try a couple examples with a new dataset

```{r}
week_8_2 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/week_8_2.csv")
```

We'll start with class differences in how respondents think people "get ahead". We can check the GSS website for a description of this variable: https://gssdataexplorer.norc.org/variables/585/vshow


## Exercise 1

At the .05 level, test the difference between the proportion of working class respondents who say that hard work is how people get ahead and the proportion of middle class respondents who say that hard work is how people get ahead. 

Start with a proportion table with `class` in the rows and `getahead` in the columns:

### REPLACE THIS LINE WITH YOUR CODE

```{r proportion table}
round(prop.table(table(week_8_2$class, 
      week_8_2$getahead),1),3)
```

```{r exercise set up, warning = FALSE, message = FALSE}

# Create Data Frame:
example1 <- week_8_2  |>
   mutate(hardwork = ifelse(getahead == "Hard work", 1, 0)) |>
   filter(class == "Working class" | class == "Middle class") |>
   droplevels()
```

This also works if you want to use str_detect with multiple strings!
```{r}
example1 <- week_8_2 |> 
   mutate(hardwork = ifelse(getahead == "Hard work", 1, 0)) |> 
   filter(str_detect(class, "Working|Middle")) |> # No space before or after |
   droplevels()
```


Now save the table and run the test.

```{r}
# Save Table:
example1_table <- table(example1$class, example1$hardwork)

# Run Test:
prop.test(example1_table)
```


## Exercise 2

Let's see if the proportions believing people get ahead through only hard work differ between Black and White respondents at the .05 alpha level.

### REPLACE THIS LINE WITH YOUR CODE

```{r}
prop.table(table(week_8_2$racehisp, week_8_2$getahead),1)
```


```{r set up binary variable and filter}
exercise2 <- week_8_2 |>
   mutate(hardwork = 
             ifelse(getahead == "Hard work", 1, 0)) |>
   filter(str_detect(racehisp, "Black|White")) |> 
   droplevels()
```

```{r proportions exercise frequency table}
exercise2_table <- table(exercise2$racehisp, exercise2$hardwork)
exercise2_table
```

And now feed this saved table into the `prop.test()` function:

```{r prop test with frequency table}
prop.test(exercise2_table)
```