## Setting Up

Load the `cupid.csv` file as a data frame called `cupid` and load the usual packages:

```{r, include = FALSE, warning = FALSE}
cupid <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385/master/data/cupid.csv")
library(tidyverse)
library(pander)
```


## Recapping Wednesday

When a variable is normally distributed, useful to know how far a value is from the mean in standardized units.

The formula for standardizing the value of a normally distributed variable is: 

$\Large{z = \frac{x-\mu}{\sigma} = \frac{x - mean}{sd}}$


Find the z-score for your height (in inches) using the `height` variable:

###REPLACE THIS LINE WITH YOUR CODE###


Find the probability that someone in this sample will be shorter than you:

###REPLACE THIS LINE WITH YOUR CODE###

Find the probability that someone in this sample will be taller than you:

###REPLACE THIS LINE WITH YOUR CODE###


## Standard Errors

The standard deviation of the sampling distribution is called the *standard error*. It is calculated as:

$\Large{\sigma_{\bar{y}} = \frac{\sigma}{\sqrt{n}} = \frac{sd}{\sqrt{sample size}}}$


Let's find the standard error of the `age` variable. We'll save this as an object, not as a new variable (since it is the same for the entire sample):

```{r}
age_se <- sd(cupid$age) / sqrt(length(cupid$age))

age_se
```


## From Standard Errors to Confidence Intervals

For a 95% confidence interval, we need the z-scores that are associated with .025 and .975. To find them, use `qnorm()`:

```{r}
qnorm(.025)

qnorm(.975)
```

In common practice, we round this to 1.96. We'll use this number a lot; remember it so you don't have to use `qnorm()` every time you need it.


## Margin of Error

The margin of error is the z-score associated with the confidence interval we are constructing multipled by the standard error:

```{r}
1.96*age_se
```


## Building Confidence Intervals

The sample mean plus and minus the margin of error is the confidence interval.

For the *lower limit* of the confidence interval:

```{r}
age_ll <- mean(cupid$age) - 1.96*age_se
```

For the *upper limit* of the confidence interval:

```{r}
age_ul <- mean(cupid$age) + 1.96*age_se
```

Save the 95% confidence interval in a vector:

```{r}
age_ci <- c(age_ll, mean(cupid$age), age_ul)

age_ci
```

How do you interpret this confidence interval? 95% of the repeated samples we might imagine pulling would be expected to have means within this range, giving us 95% confidence that the true population mean falls within this range.


## Exercise

What is the 99% confidence interval for height?

### REPLACE THIS LINE WITH YOUR CODE

What is the 90% confidence interval for age?

### REPLACE THIS LINE WITH YOUR CODE
