---
title: "Problem Set 2 Solutions"
output:
  pdf_document: default
---

```{r, message = FALSE, warning = FALSE, include = FALSE}
library(tidyverse)
library(vcdExtra)
library(broom)
ps2 <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/ps2.csv")
```

**1. Without using any R shortcuts, find the 95% confidence interval for the mean of `eqwlth` in each of the following years: 2010, 2014, and 2018. Plot these intervals in a figure (with error bars), and use your figure to describe how the mean responses have changed over the survey years.**

Students will probably do this separately for observations from each year. 

For example, from 2010:

```{r}
year_2010 <- filter(ps2, year == 2010)
mean_2010 <- mean(year_2010$eqwlth)
se_2010 <- sd(year_2010$eqwlth) / sqrt(length(year_2010$eqwlth))
ll_2010 <- mean_2010 - 1.96*se_2010
ul_2010 <- mean_2010 + 1.96*se_2010
ci_2010 <- c(ll_2010, mean_2010, ul_2010)
ci_2010
```

And from 2014:
```{r}
year_2014 <- filter(ps2, year == 2014)
mean_2014 <- mean(year_2014$eqwlth)
se_2014 <- sd(year_2014$eqwlth) / sqrt(length(year_2014$eqwlth))
ll_2014 <- mean_2014 - 1.96*se_2014
ul_2014 <- mean_2014 + 1.96*se_2014
ci_2014 <- c(ll_2014, mean_2014, ul_2014)
ci_2014
```

And from 2018:
```{r}
year_2018 <- filter(ps2, year == 2018)
mean_2018 <- mean(year_2018$eqwlth)
se_2018 <- sd(year_2018$eqwlth) / sqrt(length(year_2018$eqwlth))
ll_2018 <- mean_2018 - 1.96*se_2018
ul_2018 <- mean_2018 + 1.96*se_2018
ci_2018 <- c(ll_2018, mean_2018, ul_2018)
ci_2018
```

Once you have a confidence interval for each year, combine them into a table:

```{r}
ci_years <- round(rbind(ci_2010, ci_2014, ci_2018),3)
ci_years
```

Then clean up the table

```{r}
ci_years_rownames <- as.factor(c(2010, 2014, 2018))
ci_years_df <- cbind.data.frame(ci_years_rownames, ci_years) 
colnames(ci_years_df) <- c("Year", "LL", "Mean", "UL")
ci_years_df
```

And use the new data frame for your plot

```{r}
years_plot <- ggplot(ci_years_df, aes(x = Year, y = Mean,
                                      ymin = LL, ymax = UL))
years_plot + geom_point() + geom_errorbar() +
     labs(title = "Should Government Reduce Income Differences?",
          subtitle = "General Social Survey (2010, 2014, 2018)",
          caption = "Note: Lower values indicate more support for reducing income differences") + # not sure we have seen caption in class
  theme(plot.caption = element_text(hjust = 0))
```

Here's a shortcut using group_by and summarize:

```{r}

q1 <- ps2 |> 
  filter(year == 2010 | year == 2014 | year == 2018) |> 
  group_by(year) %>% 
     summarize(mean = mean(eqwlth),
          sd = sd(eqwlth),
          n = length(eqwlth),
          se = sd / sqrt(n),
          ll = mean - 1.96*se,
          ul = mean + 1.96*se)
q1
```

And the plot (note that with this approach you need to assert the `year` variable as a factor variable):

```{r}
plot_q1 <- ggplot(q1, aes(x = factor(year), y = mean,
                          ymin = ll, ymax = ul))

plot_q1 + geom_point() + geom_errorbar() +
     labs(x = "Year", y = "Mean",
          title = "Should Government Reduce Income Differences?",
          subtitle = "General Social Survey (2010, 2014, 2018)",
          caption = "Note: Lower values indicate more support for reducing income differences") + theme(plot.caption = element_text(hjust = 0))
```


**2. Create a new variable grouping the `age` variable into the following categories: 18-24, 25-39, 40-54, 55-64, 65+. Which (if any) age categories showed significant differences in mean `eqwlth` scores between the 2010 and 2018 surveys?**

New variable using mutate and ifelse:

```{r}
ps2 <- ps2 %>%
     mutate(agecat = 
          ifelse(age %in% 18:24, 1,
             ifelse(age %in% 25:39, 2,
               ifelse(age %in% 40:54, 3,
                 ifelse(age %in% 55:64, 4, 5)))))
```

Tests for each age category:

```{r}
t.test(ps2$eqwlth[ps2$agecat==1&ps2$year==2010],
       ps2$eqwlth[ps2$agecat==1&ps2$year==2018])

t.test(ps2$eqwlth[ps2$agecat==2&ps2$year==2010],
       ps2$eqwlth[ps2$agecat==2&ps2$year==2018])

t.test(ps2$eqwlth[ps2$agecat==3&ps2$year==2010],
       ps2$eqwlth[ps2$agecat==3&ps2$year==2018])

t.test(ps2$eqwlth[ps2$agecat==4&ps2$year==2010],
       ps2$eqwlth[ps2$agecat==4&ps2$year==2018])

t.test(ps2$eqwlth[ps2$agecat==5&ps2$year==2010],
       ps2$eqwlth[ps2$agecat==5&ps2$year==2018])
```

Can see there are significant differences in means for groups 2, 3, and 5: p-values are less than .05, test statistics are more extreme than 1.96, and zero is not in the confidence intervals.

We could also do this much more efficiently using group_by and summarise_each (but we have not seen how to do so in class so no expectation that you know how to do this).

```{r}
multiple_ttests <- ps2 |> 
  group_by(agecat) |> 
  filter(year == 2010 | year == 2018) %>%
  summarise_each(list(~t.test(.[year == 2010], .[year == 2018])$statistic,
                      ~t.test(.[year == 2010], .[year == 2018])$p.value,
                      ~t.test(.[year == 2010], .[year == 2018])$estimate[[1]],
                      ~t.test(.[year == 2010], .[year == 2018])$estimate[[2]],
                      ~t.test(.[year == 2010], .[year == 2018])$conf.int[[1]],
                      ~t.test(.[year == 2010], .[year == 2018])$conf.int[[2]]), 
                 vars = eqwlth) |> 
  mutate(difference = `vars_[[..3` - `vars_[[..4`) |> 
  select(-c(4:5))
round(multiple_ttests,3)
```


**3. Does the proportion of respondents with "Hardly any" confidence in congress differ between respondents at the lowest and highest extremes of the `eqwlth` scale?**

```{r}
# Create a binary variable
q3 <- ps2 |>  
  mutate(conf_hi = ifelse(conlegis == "Hardly any", 1, 0)) |> 
  filter(eqwlth==1 | eqwlth==7)

# Create a table
q3_table <- table(q3$eqwlth, q3$conf_hi)
round(prop.table(q3_table,1),3)

# Run prop.test on the table
prop.test(q3_table)
```

Yes, the proportions are significantly different. We can reject the null hypothesis that the difference is zero because the p-value is less than .05 and zero is not in the confidence interval. Don't look at the x-squared statistic for the proportion test.

**4a. Is there an association between racehisp and eqwlth?**

```{r}
chisq.test(ps2$racehisp, ps2$eqwlth)
```

Use a chi-squared test here because both variables are categorical but at least one (racehisp) is not ordered. Yes, there is a significant association because we can reject the null hypothesis since the p-value is less than .05. That means the two variables are dependent.


**4b. Among respondents with less than a high school diploma, is there an association between race/Hispanic status and eqwlth?**

```{r}
q4b <- filter(ps2, degree=="Less than HS")
chisq.test(q4b$racehisp, q4b$eqwlth)
```

We get an error message, so let's see if the expected frequency in each cell is at least 5.

```{r}
chisq.test(q4b$racehisp, q4b$eqwlth)$expected
```

Some cells have fewer than five expected observations, so we need to use Fisher's Test

```{r}
fisher.test(q4b$racehisp, q4b$eqwlth, simulate.p.value = TRUE)
```

Cannot reject the null hypothesis because the p-value is greater than .05. That means there is not a significant association between these variables.


**4c. Is there an association between age (use categories) and confidence in Congress?**

These variables are ordered, so use the GK Gamma test

```{r}
library(vcdExtra)

# Put conlegis in the right order
ps2$conlegis <- factor(ps2$conlegis, 
     levels = c("Hardly any", "Only some", "A great deal"))

# Make a table
q4c_table <- table(ps2$agecat, ps2$conlegis)

# Run the test on the table
GKgamma(q4c_table)
```

Then find the test statistic:
```{r}
-.215 / .015
```

There is a significant negative association: respondents in the youngest age categories tend to have more confidence in congress than respondents in the older age categories.