---
title: "Problem Set 1"
output:
  pdf_document: default
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
ps1 <- read.csv("http://raw.githubusercontent.com/mjclawrence/soci385_f21/main/data/ps1.csv")
library(tidyverse)
library(kableExtra)
```

1. In a paragraph, describe the responses to the `racespeech` question. Your paragraph should include:
- The overall proportion of respondents who agree and disagree
- The proportion of respondents in each race/hisp category who agree and disagree
- The proportion of respondents in each degree category who agree and disagree

```{r}

# Proportion table
kbl(round(prop.table(table(ps1$racespeech)),3),
    booktabs = TRUE,
    align = rep("c", 2))

# Proportion table with row proportions
kbl(round(prop.table(table(ps1$racehisp, ps1$racespeech),1),3),
    booktabs = TRUE,
    align = rep("c", 2))

# Relevel degree categories
ps1$degree <- factor(ps1$degree, 
                     levels=c("Less Than HS", "HS Diploma", 
                              "Some College", "College Degree",
                              "Grad/Prof Degree"))

# Proportion table with row proportions
kbl(round(prop.table(table(ps1$degree, ps1$racespeech),1),3),
    booktabs = TRUE,
    align = rep("c", 2))
```

2. In a figure, describe how the age distribution differs for respondents who do and do not agree with the question of whether a racist should be allowed to speak. Interpret your figure in words, and offer a possible sociological explanation for the results.

```{r}
# Density plot
racespeech_age <- ggplot(ps1, aes(x = age, color = racespeech))

racespeech_age + geom_density() + 
     labs(x = "Age", y = "Density",
          title = "Support for Racist Speech by Age",
          subtitle = "GSS, 2014",
          color = "Racist Speaker\nShould Be:")
```



3. In a sentence or two, describe male-female differences in rally attendance. What is an additional variable you would want to have in order to explain the relationship between these variables?

```{r}

# Reorder rally levels
ps1$attrally <- factor(ps1$attrally, 
                       levels = c("In Past Year", "Before Last Year",
                       "No But I Might", "No And I Will Not"))

# Proportion table with row proportions
kbl(round(prop.table(table(ps1$sex, ps1$attrally),1),3),
    booktabs = TRUE,
    align = rep("c", 4))
```


4. In a few sentences, describe class differences in demonstration participation. What is an additional variable you would want to have in order to explain the relationship between these variables?

```{r}

# Reorder class levels
ps1$class <- factor(ps1$class, 
                    levels = c("Lower", "Working", "Middle", "Upper"))

# Reorder demonstration levels
ps1$joindem <- factor(ps1$joindem, 
                      levels = c("In Past Year", "Before Last Year",
                       "No But I Might", "No And I Will Not"))

# Proportion table with row proportions
kbl(round(prop.table(table(ps1$class, ps1$joindem),1),3),
    booktabs = TRUE,
    align = rep("c", 4))
```

5. Collapse the seven categories of political views into three categories: any liberal, moderate, any conservative. Create a figure showing how the average number of political actions varies across the three categories of political views. Interpret this figure in a few sentences.

```{r}

# Create categories
ps1 <- ps1 |>
     mutate(polcat = ifelse(polviews %in% 1:3, "Liberal",
                            ifelse(polviews == 4, "Moderate",
                                   "Conservative")),
            polcat = factor(polcat,
                            levels = c(
                                 "Liberal", "Moderate", "Conservative")))

# Find mean of actions by category
polcat_action <- ps1 |>
               group_by(polcat) |>
               summarize(avg = round(mean(num_action),2))
```

\pagebreak

```{r}
# Plot
action_by_polcat <- ggplot(polcat_action, aes(x = polcat, y = avg))
action_by_polcat + geom_col() + 
     geom_text(aes(label = avg), vjust = "outward") +
     labs(x = "Political Views", 
          y = "Average Number of Political Actions",
          title = "Average Number of Political Actions by Political Views",
          subtitle = "GSS, 2014")
```

6. In a sentence or two, describe the correlation between years of education and number of political actions. (No figure required here.)

```{r}
cor(ps1$num_action, ps1$educ)
```